---
title: "Home Credit Default Risk Exploratory Data Analysis"
subtitle: "From business framing to defensible data preparation"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-overflow: wrap
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    css: styles.css
    fig-width: 9
    fig-height: 5.2
    fig-format: retina
execute:
  echo: true
  warning: false
  message: false
  error: false
  cache: false
editor: source
---

## Executive summary

This exploratory analysis finds a useful but challenging application-level modeling table. The training data contain 307,511 unique applications and 121 candidate predictors; only 8.07% of applicants experienced the payment difficulties identified by `TARGET`. External risk scores and age show the clearest individual associations with the outcome, but no single variable is sufficient for an approval decision. The data also contain substantial, patterned missingness, an 18% sentinel code in employment duration, rare invalid labels, extreme monetary observations, and a material difference in loan-type mix between the training and scoring cohorts.

These findings change the modeling plan. Preprocessing must be learned on training folds only, missingness indicators should be tested rather than treating all missing values as noise, raw time variables should be converted to interpretable units, and evaluation should emphasize ROC AUC alongside precision-recall and threshold-based business tradeoffs. Group-level rates in this notebook are descriptive associations—not causal effects or instructions to approve or deny a person.

## Business problem and EDA questions

Home Credit serves applicants who may have limited or nontraditional credit histories. At application time, the business must balance two costly errors: approving a loan that later encounters payment difficulties, and declining an applicant who would have repaid. The proposed analytic product is therefore a risk-ranking model that supports underwriters; it should not replace policy, affordability checks, or human review.

The supplied data define `TARGET = 1` as a client who had sufficiently late payment on at least one of the first installments, with the precise lateness and installment thresholds anonymized in the data dictionary. Model discrimination can be compared using ROC AUC, but deployment success should ultimately be judged at a decision threshold: default or difficulty rate at a given approval volume, approval volume at a risk limit, calibration, and performance across relevant groups.

EDA bridges that framing to data preparation by answering five questions:

1. What is the observation grain, and can the training and test files be combined or joined safely?
2. How imbalanced is the target, and what does that imply for evaluation?
3. Which values are missing, invalid, sentinel-coded, or unusually extreme?
4. Which applicant and loan characteristics have the strongest descriptive relationship with payment difficulty?
5. Are the training and scoring cohorts similar enough for a model trained on one to generalize to the other?

## Setup and data loading

Only `application_train.csv` and `application_test.csv` are analyzed, as required for this stage. The supplementary relational tables are deliberately left out so that the application-level grain remains explicit.

```{r}
# Load only the packages needed for manipulation, plots, and compact tables.
library(data.table)
library(ggplot2)
library(scales)
library(knitr)

# Make paths work whether Quarto executes from the project root or notebook folder.
project_root <- if (file.exists("data/raw/application_train.csv")) "." else ".."
train_path <- file.path(project_root, "data", "raw", "application_train.csv")
test_path  <- file.path(project_root, "data", "raw", "application_test.csv")
dictionary_path <- file.path(project_root, "docs", "HomeCredit_columns_description.csv")

# Treat blank fields as missing and suppress the progress bar in the rendered report.
train <- fread(train_path, na.strings = c("", "NA"), showProgress = FALSE)
test  <- fread(test_path,  na.strings = c("", "NA"), showProgress = FALSE)
dictionary <- fread(dictionary_path, encoding = "Latin-1", showProgress = FALSE)

# Use a restrained, consistent visual style throughout the notebook.
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title.position = "plot",
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "#52606d"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)

navy <- "#16324F"
teal <- "#168AAD"
coral <- "#D95D39"
gold <- "#D9A441"

# Reusable helpers keep formatting decisions out of the analytical code below.
pct <- function(x, accuracy = 0.1) percent(x, accuracy = accuracy)
fmt_n <- function(x) comma(x, accuracy = 1)
```

The files are read without changing the raw source. All derived variables exist only in memory, which keeps the audit trail reproducible.

## Structure, grain, and keys

```{r}
# Summarize table shape, key uniqueness, and target availability.
structure_summary <- data.table(
  Dataset = c("Training applications", "Scoring applications"),
  Rows = c(nrow(train), nrow(test)),
  Columns = c(ncol(train), ncol(test)),
  `Unique SK_ID_CURR` = c(uniqueN(train$SK_ID_CURR), uniqueN(test$SK_ID_CURR)),
  `Duplicate keys` = c(sum(duplicated(train$SK_ID_CURR)), sum(duplicated(test$SK_ID_CURR))),
  `TARGET present` = c("Yes", "No")
)

kable(structure_summary, format.args = list(big.mark = ","), align = "lrrrrc",
      caption = "Application table structure and key audit")
```

Both files have exactly one row per `SK_ID_CURR`; there are no duplicate keys and no identifier overlap across files. The training file has one extra column because `TARGET` is withheld from the scoring set. Thus, these tables should be **row-bound only for consistent preprocessing or population comparisons**, never joined to one another. If supplementary tables are later added, their many-to-one or many-to-many histories must first be aggregated to `SK_ID_CURR`; a direct join would duplicate application rows and distort both rates and model weights.

```{r}
# Count storage types to understand the mix of continuous, count, flag, and label fields.
type_counts <- data.table(
  Storage_type = names(table(vapply(train, function(x) class(x)[1], character(1)))),
  Columns = as.integer(table(vapply(train, function(x) class(x)[1], character(1))))
)
kable(type_counts, caption = "Training columns by imported storage type")
```

Storage type is not semantic type. Many imported integers are binary flags or ordered ratings, while some numeric fields are normalized scores. This matters because blindly scaling all numeric columns or treating all integers as continuous would discard useful meaning.

## Target balance and evaluation implications

```{r}
# Calculate class counts and label them in business language.
target_summary <- train[, .(Applications = .N), by = TARGET]
target_summary[, Share := Applications / sum(Applications)]
target_summary[, Outcome := fifelse(TARGET == 1,
                                    "Payment difficulties", "No payment difficulties")]

ggplot(target_summary, aes(x = Outcome, y = Applications, fill = factor(TARGET))) +
  geom_col(width = 0.62, show.legend = FALSE) +
  geom_text(aes(label = paste0(fmt_n(Applications), "\n", pct(Share, 0.01))),
            vjust = -0.35, fontface = "bold", size = 3.8) +
  scale_fill_manual(values = c(navy, coral)) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.14))) +
  labs(
    title = "Payment difficulties occur in 8.07% of training applications",
    subtitle = "A majority-class prediction would be 91.93% accurate but operationally useless",
    x = NULL, y = "Number of applications"
  )
```

The 11.4-to-1 class imbalance makes accuracy a misleading headline metric. ROC AUC matches the project's ranking goal, but precision-recall AUC is also needed because it focuses attention on the minority class. Final model selection should add calibration and threshold tables showing approvals, flagged applicants, and observed difficulty rates. Stratified cross-validation is appropriate; resampling or class weights, if used, must occur within training folds.

## Missing values

### Extent and concentration

```{r}
# Compute missingness for every field, then display the most incomplete features.
missing_profile <- data.table(
  Feature = names(train),
  Missing = vapply(train, function(x) sum(is.na(x)), integer(1)),
  Missing_rate = vapply(train, function(x) mean(is.na(x)), numeric(1))
)
missing_profile <- missing_profile[order(-Missing_rate)]

top_missing <- missing_profile[1:20]
ggplot(top_missing, aes(x = reorder(Feature, Missing_rate), y = Missing_rate)) +
  geom_col(fill = teal, width = 0.72) +
  geom_text(aes(label = pct(Missing_rate)), hjust = -0.12, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.78)) +
  labs(
    title = "Missingness is concentrated in housing and asset fields",
    subtitle = paste0(sum(missing_profile$Missing_rate > 0.50),
                      " features are more than 50% missing; ",
                      sum(missing_profile$Missing_rate == 0), " are complete"),
    x = NULL, y = "Share missing"
  )
```

The highest-missing fields are repeated measurements of common area, non-living apartments, living apartments, minimum floors, building age, and other housing characteristics. Their shared patterns suggest that missingness often reflects whether a property record or applicable asset exists, not random data loss. For example, `OWN_CAR_AGE` is absent for most non-car owners. Automatic column deletion at a single threshold would erase this structure.

The three external scores differ sharply in coverage: `EXT_SOURCE_1` is 56.38% missing, `EXT_SOURCE_3` is 19.83% missing, and `EXT_SOURCE_2` is only 0.21% missing. Since the scores later show strong outcome separation, imputation and missingness indicators should be validated carefully.

### Missingness may itself be informative

```{r}
# Compare missing rates by target and rank features by the absolute gap.
missing_by_target <- rbindlist(lapply(names(train), function(v) {
  data.table(
    Feature = v,
    `No difficulties` = mean(is.na(train[TARGET == 0][[v]])),
    `Payment difficulties` = mean(is.na(train[TARGET == 1][[v]]))
  )
}))
missing_by_target[, Gap := `Payment difficulties` - `No difficulties`]
missing_gap <- missing_by_target[order(-abs(Gap))][1:12]

missing_gap_long <- melt(
  missing_gap,
  id.vars = c("Feature", "Gap"),
  measure.vars = c("No difficulties", "Payment difficulties"),
  variable.name = "Outcome", value.name = "Missing_rate"
)

ggplot(missing_gap_long,
       aes(x = reorder(Feature, abs(Gap)), y = Missing_rate, color = Outcome)) +
  geom_point(size = 2.8) +
  geom_line(aes(group = Feature), color = "#b8c2cc", linewidth = 0.8) +
  coord_flip() +
  scale_color_manual(values = c(navy, coral)) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Property-data absence is more common among difficult-payment cases",
    subtitle = "Twelve features with the largest target-group missingness gaps",
    x = NULL, y = "Share missing", color = "Observed outcome"
  )
```

For several property fields, the missing rate is roughly seven percentage points higher among applicants with payment difficulties. This does not prove that missing property data causes difficulty: missingness can proxy for housing type, wealth, geography, or the data collection process. It does justify testing explicit missing-value indicators and comparing them with models that handle missingness natively. All imputation values and missing-indicator decisions must be learned inside cross-validation to prevent leakage.

## Validity, sentinel codes, and unusual values

```{r}
# Quantify data-quality conditions that require an explicit preparation decision.
quality_issues <- data.table(
  Check = c(
    "DAYS_EMPLOYED equals 365,243",
    "CODE_GENDER equals XNA",
    "NAME_FAMILY_STATUS equals Unknown",
    "CNT_FAM_MEMBERS is missing",
    "AMT_ANNUITY is missing",
    "AMT_GOODS_PRICE is missing",
    "AMT_INCOME_TOTAL exceeds 10 million",
    "OWN_CAR_AGE exceeds 60 years"
  ),
  Rows = c(
    sum(train$DAYS_EMPLOYED == 365243, na.rm = TRUE),
    sum(train$CODE_GENDER == "XNA", na.rm = TRUE),
    sum(train$NAME_FAMILY_STATUS == "Unknown", na.rm = TRUE),
    sum(is.na(train$CNT_FAM_MEMBERS)),
    sum(is.na(train$AMT_ANNUITY)),
    sum(is.na(train$AMT_GOODS_PRICE)),
    sum(train$AMT_INCOME_TOTAL > 1e7, na.rm = TRUE),
    sum(train$OWN_CAR_AGE > 60, na.rm = TRUE)
  )
)
quality_issues[, Share := Rows / nrow(train)]
kable(quality_issues, digits = 4, format.args = list(big.mark = ","),
      caption = "Selected validity and outlier checks")
```

`DAYS_BIRTH`, `DAYS_REGISTRATION`, and `DAYS_ID_PUBLISH` are non-positive as documented, but `DAYS_EMPLOYED = 365243` is about 1,000 years and cannot be a real tenure. It affects 55,374 rows (18.01%) and should be recoded to missing with a separate sentinel indicator before converting valid negative values to positive years. This is safer than allowing a model to interpret the code as an enormous continuous value.

The four `XNA` gender labels and two `Unknown` family-status labels should be retained as explicit unknown categories or recoded consistently—not guessed. Extreme incomes and old cars are suspicious but not automatically errors; robust transforms, winsorization determined on training folds, and sensitivity checks are preferable to deleting applicants. There are no nonpositive income or credit amounts, no applicants with more children than family members, and valid negative employment durations do not exceed applicant age.

```{r}
# Create interpretable analysis variables without overwriting source columns.
eda <- copy(train)
eda[, AGE_YEARS := -DAYS_BIRTH / 365.25]
eda[, EMPLOYED_SENTINEL := DAYS_EMPLOYED == 365243]
eda[, EMPLOYED_YEARS := fifelse(DAYS_EMPLOYED < 0, -DAYS_EMPLOYED / 365.25, NA_real_)]
eda[, CREDIT_INCOME_RATIO := AMT_CREDIT / AMT_INCOME_TOTAL]
eda[, ANNUITY_INCOME_RATIO := AMT_ANNUITY / AMT_INCOME_TOTAL]
eda[, CREDIT_GOODS_RATIO := AMT_CREDIT / AMT_GOODS_PRICE]
eda[, EXTERNAL_SCORE_COUNT := rowSums(!is.na(.SD)),
    .SDcols = c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")]
```

## Applicant profile and outcome relationships

### Age

```{r}
# Bin age into interpretable decades and show both volume and outcome rate.
age_summary <- eda[, .(
  Applications = .N,
  Difficulty_rate = mean(TARGET)
), by = .(Age_band = cut(
  AGE_YEARS, breaks = c(20, 30, 40, 50, 60, 70),
  right = FALSE, include.lowest = TRUE
))][order(Age_band)]

ggplot(age_summary, aes(x = Age_band, y = Difficulty_rate)) +
  geom_col(aes(fill = Applications), width = 0.7) +
  geom_text(aes(label = paste0(pct(Difficulty_rate, 0.01), "\nn=", fmt_n(Applications))),
            vjust = -0.35, size = 3.4) +
  scale_fill_gradient(low = "#8ecae6", high = navy, guide = "none") +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.135)) +
  labs(
    title = "Observed payment difficulty falls steadily with applicant age",
    subtitle = "Rates are descriptive and may reflect correlated income, employment, or product mix",
    x = "Age at application (years)", y = "Observed difficulty rate",
    fill = "Applicants"
  )
```

The observed rate declines from 11.44% among applicants aged 20–29 to 4.92% among those aged 60–69. The monotonic pattern makes age potentially predictive, but it also creates fairness and policy concerns. Age should not be interpreted causally, and model performance and decisions should be audited by age band before deployment.

### External source scores

```{r}
# Use within-feature quintiles so the three differently distributed scores are comparable.
external_long <- rbindlist(lapply(
  c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3"),
  function(v) {
    x <- eda[!is.na(get(v)), .(Score = get(v), TARGET)]
    breaks <- unique(quantile(x$Score, probs = seq(0, 1, 0.2), na.rm = TRUE))
    x[, Quintile := cut(Score, breaks = breaks, include.lowest = TRUE, labels = FALSE)]
    x[, .(Applications = .N, Difficulty_rate = mean(TARGET)), by = Quintile][
      , Source := v]
  }
))

ggplot(external_long,
       aes(x = factor(Quintile), y = Difficulty_rate, color = Source, group = Source)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c(navy, teal, gold),
                     labels = c("External source 1", "External source 2", "External source 3")) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Higher external scores consistently correspond to lower observed risk",
    subtitle = "Quintile 1 contains the lowest scores and quintile 5 the highest",
    x = "Within-score quintile", y = "Observed difficulty rate", color = NULL
  )
```

The external scores provide the strongest single-variable separation in this table. For `EXT_SOURCE_3`, the difficulty rate falls from about 16.38% in the lowest-score quintile to 3.38% in the highest; `EXT_SOURCE_1` and `EXT_SOURCE_2` show similarly ordered patterns. Because the scores are anonymized, their provenance, permissible use, stability, and overlap with protected or sensitive information need governance review. Missingness also prevents treating their observed subsets as the entire population.

```{r}
# Measure whether merely having more external scores is associated with the outcome.
score_coverage <- eda[, .(
  Applications = .N,
  Difficulty_rate = mean(TARGET)
), by = EXTERNAL_SCORE_COUNT][order(EXTERNAL_SCORE_COUNT)]

kable(score_coverage, digits = 4, format.args = list(big.mark = ","),
      col.names = c("External scores available", "Applications", "Difficulty rate"),
      caption = "External score coverage and observed payment difficulty")
```

Applicants with all three scores available have a lower observed difficulty rate (7.30%) than those with only one (9.92%). The group with no scores is very small (172), so its 8.14% rate is unstable. Availability is potentially useful information, but it may encode access to external data rather than underlying creditworthiness.

### Selected categorical characteristics

```{r}
# Compare sufficiently large categories; small groups can produce unstable rates.
category_rate <- function(variable, min_n = 1000) {
  eda[, .(Applications = .N, Difficulty_rate = mean(TARGET)), by = variable][
    Applications >= min_n][order(Difficulty_rate)]
}

housing <- category_rate("NAME_HOUSING_TYPE")
ggplot(housing, aes(x = reorder(NAME_HOUSING_TYPE, Difficulty_rate),
                    y = Difficulty_rate)) +
  geom_col(fill = teal, width = 0.7) +
  geom_text(aes(label = paste0(pct(Difficulty_rate, 0.1), "  n=", fmt_n(Applications))),
            hjust = -0.08, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.14)) +
  labs(
    title = "Housing categories differ in observed payment difficulty",
    subtitle = "Only categories with at least 1,000 training applications are shown",
    x = NULL, y = "Observed difficulty rate"
  )
```

Applicants in rented housing (12.31%) or living with parents (11.70%) have higher observed rates than those in a house or apartment (7.80%). This relationship may be useful for prediction but is intertwined with age, income, geography, and loan type. Multivariable analysis is required, and policy owners should review whether and how housing variables may be used.

```{r}
# Show large education categories with their sample sizes to keep rate comparisons honest.
education <- category_rate("NAME_EDUCATION_TYPE")
ggplot(education, aes(x = reorder(NAME_EDUCATION_TYPE, Difficulty_rate),
                      y = Difficulty_rate)) +
  geom_col(fill = navy, width = 0.7) +
  geom_text(aes(label = paste0(pct(Difficulty_rate, 0.1), "  n=", fmt_n(Applications))),
            hjust = -0.08, size = 3.2) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.125)) +
  labs(
    title = "Education level is associated with observed difficulty",
    subtitle = "Only categories with at least 1,000 training applications are shown",
    x = NULL, y = "Observed difficulty rate"
  )
```

The observed rate ranges from 5.36% for higher education to 10.93% for lower secondary education among adequately sized categories. As with housing, this is association rather than cause. Education can act as a socioeconomic proxy, so fairness review and comparison with a model excluding sensitive proxies are appropriate.

### Loan size and affordability ratios

```{r}
# Trim only for visualization so a few extreme incomes do not compress the plot.
amount_plot <- melt(
  eda[, .(
    TARGET,
    `Annual income` = AMT_INCOME_TOTAL,
    `Credit amount` = AMT_CREDIT
  )],
  id.vars = "TARGET", variable.name = "Measure", value.name = "Amount"
)
amount_plot <- amount_plot[
  , .SD[Amount <= quantile(Amount, 0.99, na.rm = TRUE)], by = Measure
]
amount_plot[, Outcome := fifelse(TARGET == 1, "Payment difficulties", "No difficulties")]

ggplot(amount_plot, aes(x = Amount, fill = Outcome)) +
  geom_density(alpha = 0.38, adjust = 1.2) +
  facet_wrap(~Measure, scales = "free", ncol = 1) +
  scale_fill_manual(values = c(navy, coral)) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Income and credit distributions overlap substantially across outcomes",
    subtitle = "Display is limited to the within-measure 99th percentile; no rows are removed from analysis",
    x = "Amount", y = "Density", fill = "Observed outcome"
  )
```

Raw income and credit amount alone show extensive overlap. This suggests that relative affordability features may be more meaningful than absolute currency values, while also warning against interpreting a high income as automatically low risk.

```{r}
# Summarize engineered ratios and their simple point-biserial correlations with TARGET.
ratio_summary <- data.table(
  Feature = c("Credit / income", "Annuity / income", "Credit / goods price"),
  Median = c(
    median(eda$CREDIT_INCOME_RATIO, na.rm = TRUE),
    median(eda$ANNUITY_INCOME_RATIO, na.rm = TRUE),
    median(eda$CREDIT_GOODS_RATIO, na.rm = TRUE)
  ),
  `99th percentile` = c(
    quantile(eda$CREDIT_INCOME_RATIO, 0.99, na.rm = TRUE),
    quantile(eda$ANNUITY_INCOME_RATIO, 0.99, na.rm = TRUE),
    quantile(eda$CREDIT_GOODS_RATIO, 0.99, na.rm = TRUE)
  ),
  `Correlation with TARGET` = c(
    cor(eda$CREDIT_INCOME_RATIO, eda$TARGET, use = "complete.obs"),
    cor(eda$ANNUITY_INCOME_RATIO, eda$TARGET, use = "complete.obs"),
    cor(eda$CREDIT_GOODS_RATIO, eda$TARGET, use = "complete.obs")
  )
)
kable(ratio_summary, digits = 3, caption = "Candidate affordability features")
```

The ratio summaries are plausible feature-engineering candidates, but their linear correlations are weak. The credit-to-goods-price ratio has the clearest association of the three. A flexible model may capture nonlinearities and interactions, but each ratio needs guarded division, missing-value handling, and fold-specific treatment of extremes.

## Numeric feature screening

```{r}
# Rank numeric variables by absolute point-biserial correlation as a screening tool.
# Correlation is not feature importance and does not capture nonlinear relationships.
numeric_features <- setdiff(
  names(eda)[vapply(eda, is.numeric, logical(1))],
  c("TARGET", "SK_ID_CURR")
)
target_correlations <- rbindlist(lapply(numeric_features, function(v) {
  data.table(
    Feature = v,
    Correlation = cor(eda[[v]], eda$TARGET, use = "pairwise.complete.obs")
  )
}))
top_correlations <- target_correlations[order(-abs(Correlation))][1:15]

ggplot(top_correlations,
       aes(x = reorder(Feature, abs(Correlation)), y = Correlation,
           fill = Correlation > 0)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  coord_flip() +
  scale_fill_manual(values = c(teal, coral)) +
  labs(
    title = "External scores lead simple numeric association screening",
    subtitle = "Positive values correspond to higher observed payment difficulty",
    x = NULL, y = "Point-biserial correlation with TARGET"
  )
```

The three external scores have the largest absolute correlations (about −0.16 to −0.18). Age, regional rating, phone-change recency, and identity-document recency are smaller signals. `DAYS_EMPLOYED` appears in the ranking only after being contaminated by its sentinel; models should use the cleaned duration and sentinel indicator instead. Correlation is only a screen: it misses nonlinear patterns, interactions, and information in categorical features.

## Training and scoring population comparison

The test file is not a random holdout labeled for local evaluation; it is the separate cohort to be scored. Population differences therefore matter even when column names match.

```{r}
# Compare the most visibly shifted categorical field in the two cohorts.
contract_mix <- rbindlist(list(
  train[, .(Applications = .N), by = NAME_CONTRACT_TYPE][, Dataset := "Training"],
  test[,  .(Applications = .N), by = NAME_CONTRACT_TYPE][, Dataset := "Scoring"]
))
contract_mix[, Share := Applications / sum(Applications), by = Dataset]

ggplot(contract_mix, aes(x = Dataset, y = Share, fill = NAME_CONTRACT_TYPE)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = fifelse(Share < 0.02, "", pct(Share, 0.1))),
            position = position_stack(vjust = 0.5), color = "white", fontface = "bold") +
  scale_fill_manual(values = c(navy, gold)) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "The scoring cohort contains far fewer revolving loans",
    subtitle = "Loan-type mix shifts even though schemas align",
    x = NULL, y = "Share of applications", fill = "Contract type"
  )
```

Revolving loans account for 9.52% of training applications but only 0.90% of scoring applications. Median credit is also lower in the scoring data (450,000 versus 513,531), while median income is higher (157,500 versus 147,150). These differences can change calibration and the contribution of product-related features.

```{r}
# Quantify selected cohort differences with directly interpretable summaries.
cohort_summary <- data.table(
  Measure = c(
    "Revolving loan share",
    "Median credit amount",
    "Median annual income",
    "EXT_SOURCE_1 missing",
    "EXT_SOURCE_3 missing"
  ),
  Training = c(
    mean(train$NAME_CONTRACT_TYPE == "Revolving loans"),
    median(train$AMT_CREDIT, na.rm = TRUE),
    median(train$AMT_INCOME_TOTAL, na.rm = TRUE),
    mean(is.na(train$EXT_SOURCE_1)),
    mean(is.na(train$EXT_SOURCE_3))
  ),
  Scoring = c(
    mean(test$NAME_CONTRACT_TYPE == "Revolving loans"),
    median(test$AMT_CREDIT, na.rm = TRUE),
    median(test$AMT_INCOME_TOTAL, na.rm = TRUE),
    mean(is.na(test$EXT_SOURCE_1)),
    mean(is.na(test$EXT_SOURCE_3))
  )
)

# Build character display columns so percentages and currency can coexist safely.
cohort_display <- cohort_summary[, .(
  Measure,
  Training = c(
    pct(Training[1], 0.01),
    fmt_n(Training[2:3]),
    pct(Training[4:5], 0.01)
  ),
  Scoring = c(
    pct(Scoring[1], 0.01),
    fmt_n(Scoring[2:3]),
    pct(Scoring[4:5], 0.01)
  )
)]
kable(cohort_display, caption = "Selected training-versus-scoring cohort comparisons")
```

`EXT_SOURCE_1` is notably more complete in scoring data (42.12% missing versus 56.38% in training). These shifts do not make modeling impossible, but they argue for product-stratified validation, cohort-drift reporting, and calibration monitoring. Rare labels present only in training (`XNA`, `Unknown`, and `Maternity leave`) also require an encoder that can safely handle unseen or absent categories.

## Preparation decisions informed by EDA

| Finding | Preparation decision | Validation safeguard |
|---|---|---|
| One row per applicant and unique IDs | Keep `SK_ID_CURR` only as an identifier; do not use it as a predictor | Assert key uniqueness after every merge |
| `TARGET` exists only in training | Separate outcome before any preprocessing | Fit every transform on training folds only |
| 8.07% positive class | Use stratified resampling; consider weights in candidate models | Report ROC AUC, PR AUC, calibration, confusion matrix, and business threshold metrics |
| Structured, high missingness | Compare native missing handling with fold-specific imputation; add selected missing indicators | Audit performance by missingness pattern and cohort |
| `DAYS_EMPLOYED = 365243` sentinel | Recode to missing, retain a sentinel flag, convert valid duration to years | Test that no positive durations remain in cleaned field |
| Skewed currency values and extreme observations | Use log/robust transforms or tree models; cap only from training-fold estimates if needed | Compare capped and uncapped sensitivity results |
| Strong external-score signal with incomplete coverage | Retain scores and score-count indicators, subject to governance | Evaluate models with and without external scores |
| Train/test product and coverage shift | Add product-aware validation and drift monitoring | Report performance by contract type and recalibrate if needed |
| Socioeconomic and demographic proxies | Avoid causal claims and audit disparate errors | Compare subgroup calibration and false-positive/false-negative rates |

This plan keeps data-quality decisions visible and testable. It also avoids a common EDA failure: using the full labeled data to choose imputation, outlier, or encoding parameters before cross-validation.

## Results and implications

1. **The application tables have a clean grain but require join discipline.** Each row is a unique current application. Future one-to-many sources must be aggregated before joining, or application rows and outcome rates will be duplicated.
2. **The outcome is strongly imbalanced.** Payment difficulties occur in 24,825 of 307,511 training cases (8.07%). Accuracy is unsuitable by itself; ranking, precision-recall, calibration, and threshold economics are required.
3. **Missingness is a central feature of the data-generating process.** Forty-one fields exceed 50% missingness, particularly property fields, and missingness rates differ by outcome. Missing indicators and native missing-value models deserve comparison.
4. **Several values need explicit cleaning.** The 365,243 employment-day code is impossible and widespread; rare invalid categorical labels and extreme monetary values also require consistent treatment.
5. **External scores and age have the clearest descriptive relationships.** All three external scores separate outcome rates monotonically across quintiles, while observed difficulty decreases steadily across age bands. These variables require governance and fairness review, not just predictive evaluation.
6. **Simple loan amounts are not enough.** Income and credit distributions overlap substantially across outcomes, and straightforward affordability ratios have weak linear relationships. Nonlinear models and interactions are likely to be useful.
7. **Generalization risk is visible before modeling.** The scoring cohort has a sharply different contract mix, smaller median credit, higher median income, and different external-score coverage. Validation should mimic those shifts where possible and monitor calibration by product.

Overall, the EDA supports a classification workflow that begins with a transparent baseline (regularized logistic regression) and compares it with a missing-aware tree-boosting model. A final approach should be selected on out-of-fold discrimination, calibration, operational tradeoffs, stability across contract types, and fairness diagnostics—not on a single aggregate score.

## How AI was used

Generative AI was used as a coding and editorial assistant to help translate the business statement into EDA questions, draft reproducible R code, suggest data-quality checks, and organize the narrative. It was especially useful for quickly proposing audits for grain, missingness, sentinel values, class imbalance, and train/test drift, and for turning raw output into candidate interpretations.

The process also required active verification. Every reported count, percentage, plot, and table was produced by executing the notebook against the supplied CSV files; column meanings were checked against the provided data dictionary; and claims were narrowed when the evidence was only associational. AI suggestions that would have implied causality, treated suspicious outliers as certain errors, or recommended preprocessing outside cross-validation were not accepted. The main benefit was speed and breadth of checks; the main risk was confident-sounding interpretation without enough domain context. Reproducible execution and human review remain necessary before modeling or decision use.

## Reproducibility

```{r}
# Record the software environment so another analyst can investigate differences.
sessionInfo()
```
